On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2019-05-08 manuscripts/IEEE/final_submission.tex
clearer txt after thm 1

As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$.  Finally, we say that a set of vectors sharing support $S$ are in \emph{general linear position} when any $|S|$ of them are linearly independent.

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the regularity $r$ of $\mathcal{H}$ or the difference $\overline m - m$ between the assumed and actual number of columns in the latent dictionary, the more columns and coefficients of the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$ are guaranteed to be contained (up to noise) in the appropriately scaled recovered dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$, respectively. In the particular case when $\overline m = m$, the theorem directly implies that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed for $\varepsilon$ in Def.~\ref{maindef} given by: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

We claim that the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} are easily met with deterministic constructions. In particular, sparse codes in general linear position are straightforward to produce using a ``Vandermonde'' matrix construction (i.e. use the columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$).   % (Prob.~\ref{InverseProblem}).
%=======
%An immediate practical implication of this result is that there exists a practical procedure to affirm if one's proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities): simply check that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, a more general result (stated clearly in the next section) can be gleaned from our method of proving Thm.~\ref{DeterministicUniquenessTheorem}. Briefly, in cases where $\mathbf{B}$ has $\overline m \neq m$ columns, or $\mathcal{H}$ is not regular or only partially satisfying the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ gives indices $J \subseteq [m]$ defining a submatrix $\mathbf{A}_J$ and subvectors $\mathbf{x}_i^J$ that are recoverable in the sense of \eqref{Cstable} and \eqref{b-PDa}. For example, if $\mathcal{H}$ is $\ell$-regular with the SIP but $m \leq \overline m < m\ell/(\ell - 1)$ then we have nonzero $|J| = \overline m - \ell(\overline m - m)$. The implication here is that the smaller the difference $\overline m - m$, the more columns and code entries of the original $n \times m$ dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \overline m$ dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$. When $\overline m = m$, we recover Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, even if $\mathcal{H}$ is not regular or only partially satisfies the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ may give the indices $J \subseteq [m]$. For sake of brevity, we delay to the next section a clear statement of this more general result.
%
%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, it so happens that sparse codes $\mathbf{x}_i$ in general linear position are straightforward to produce with a ``Vandermonde'' matrix construction \cite{Hillar15}, leading to the following.
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

\begin{corollary}\label{DeterministicUniquenessCorollary}
Given any regular hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ with the SIP, there are $N =  |\mathcal{H}| \left[ (k-1){m \choose k} + 1  \right]$ vectors \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} such that every matrix $\mathbf{A}$ with $L_{2\mathcal{H}}(\mathbf{A}) > 0$ generates a dataset $Y = \{\mathbf{A}\mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$ with a stable $k$-sparse representation in $\mathbb{R}^m$ for $\varepsilon(\delta_1,\delta_2)$ as in \eqref{epsdel}.
\end{corollary}
% Needs to be L_{2k}(A) > 0 to guarantee sparse vector recovery

@@

One can also easily verify that for every $k < m$ there are regular $k$-uniform hypergraphs with the SIP besides the obvious $\mathcal{H} = {[m] \choose k}$. For instance, take $\mathcal{H}$ to be the $k$-regular set of consecutive intervals of length $k$ in some cyclic order on $[m]$. In this case, a direct consequence of Cor.~\ref{DeterministicUniquenessCorollary} is rigorous verification of the lower bound \mbox{$N = m(k-1){m \choose k} + m$} for sufficient sample size from the introduction. Special cases allow for even smaller hypergraphs, however. For example, if $k = \sqrt{m}$ then a 2-regular $k$-uniform hypergraph with the SIP can be constructed as the $2k$ rows and columns formed by arranging the elements of $[m]$ into a square grid.

%\cite{li2004analysis, Georgiev05, Aharon06, Hillar15}
We should stress the point here that framing the problem in terms of hypergraphs has allowed us to show, unlike in previous works on the subject, that the matrix $\mathbf{A}$ need not necessarily satisfy \eqref{SparkCondition} to be recoverable from data. As an example, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ with $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$ and take $\mathcal{H}$ to be all consecutive pairs of indices $1, \ldots ,6$ arranged in cyclic order. Then for $k=2$, the matrix $\mathbf{A}$ fails to satisfy \eqref{SparkCondition} while still satisfying the first assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.%, hence guaranteeing \eqref{Cstable} . %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
%This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\overline x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

>>>>

As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$.  Finally, we say that a set of vectors sharing support $S$ are in \emph{general linear position} when any $|S|$ of them are linearly independent.
Note that sparse codes in general linear position are straightforward to produce deterministically with a ``Vandermonde''  construction (i.e., by choosing columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$).

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the regularity $r$ of original support hypergraph $\mathcal{H}$ or the difference $\overline m - m$ between assumed and actual number of elements in the latent dictionary, the more columns and coefficients of the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$ are guaranteed to be contained (up to noise) in the appropriately labelled and scaled recovered dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$, respectively.

In the important special case when $\overline m = m$, the theorem directly implies that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed in Def.~\ref{maindef} for the following worst-case error $\varepsilon$: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

\begin{corollary}\label{DeterministicUniquenessCorollary}
Given any regular hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ with the SIP, there are $N =  |\mathcal{H}| \left[ (k-1){m \choose k} + 1  \right]$ vectors \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} such that every matrix $\mathbf{A}$ with $L_{2k}(\mathbf{A}) > 0$ generates a dataset $Y = \{\mathbf{A}\mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$ with a stable $k$-sparse representation in $\mathbb{R}^m$ for $\varepsilon(\delta_1,\delta_2)$ given by \eqref{epsdel}.
\end{corollary}

% prev
%\begin{corollary}\label{DeterministicUniquenessCorollary}
%Given any regular hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ with the SIP, there are $N =  |\mathcal{H}| \left[ (k-1){m \choose k} + 1  \right]$ vectors \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} such that every matrix $\mathbf{A}$ with $L_{2\mathcal{H}}(\mathbf{A}) > 0$ generates a dataset $Y = \{\mathbf{A}\mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$ with a stable $k$-sparse representation in $\mathbb{R}^m$ for $\varepsilon(\delta_1,\delta_2)$ as in \eqref{epsdel}.
%\end{corollary}

%We claim that the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} are easily met with deterministic constructions. In particular, sparse codes in general linear position are straightforward to produce using a ``Vandermonde'' matrix construction (i.e. use the columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$).   %

% (Prob.~\ref{InverseProblem}).
%=======
%An immediate practical implication of this result is that there exists a practical procedure to affirm if one's proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities): simply check that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, a more general result (stated clearly in the next section) can be gleaned from our method of proving Thm.~\ref{DeterministicUniquenessTheorem}. Briefly, in cases where $\mathbf{B}$ has $\overline m \neq m$ columns, or $\mathcal{H}$ is not regular or only partially satisfying the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ gives indices $J \subseteq [m]$ defining a submatrix $\mathbf{A}_J$ and subvectors $\mathbf{x}_i^J$ that are recoverable in the sense of \eqref{Cstable} and \eqref{b-PDa}. For example, if $\mathcal{H}$ is $\ell$-regular with the SIP but $m \leq \overline m < m\ell/(\ell - 1)$ then we have nonzero $|J| = \overline m - \ell(\overline m - m)$. The implication here is that the smaller the difference $\overline m - m$, the more columns and code entries of the original $n \times m$ dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \overline m$ dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$. When $\overline m = m$, we recover Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, even if $\mathcal{H}$ is not regular or only partially satisfies the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ may give the indices $J \subseteq [m]$. For sake of brevity, we delay to the next section a clear statement of this more general result.
%
%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, it so happens that sparse codes $\mathbf{x}_i$ in general linear position are straightforward to produce with a ``Vandermonde'' matrix construction \cite{Hillar15}, leading to the following.
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

% Needs to be L_{2k}(A) > 0 to guarantee sparse vector recovery

@@

One can easily verify that for every $k < m$ there are regular $k$-uniform hypergraphs $\mathcal{H}$ with the SIP besides the obvious $\mathcal{H} = {[m] \choose k}$. For instance, take $\mathcal{H}$ to be the $k$-regular set of consecutive intervals of length $k$ in some cyclic order on $[m]$. In this case, a direct consequence of Cor.~\ref{DeterministicUniquenessCorollary} is rigorous verification of the lower bound \mbox{$N = m(k-1){m \choose k} + m$} for sufficient sample size from the introduction. Special cases allow for even smaller hypergraphs, however. For example, if $k = \sqrt{m}$, then a 2-regular $k$-uniform hypergraph with the SIP can be constructed as the $2k$ rows and columns formed by arranging the elements of $[m]$ into a square grid.

%\cite{li2004analysis, Georgiev05, Aharon06, Hillar15}
We should stress the point here that framing the problem in terms of hypergraphs has allowed us to show, unlike in previous works on the subject, that the matrix $\mathbf{A}$ need not necessarily satisfy \eqref{SparkCondition} to be recoverable from data. As an example, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ with $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$ and take $\mathcal{H}$ to be all consecutive pairs of indices $1, \ldots ,6$ arranged in cyclic order. Then for $k=2$, the matrix $\mathbf{A}$ fails to satisfy \eqref{SparkCondition} while still satisfying the first set of assumptions in Thm.~\ref{DeterministicUniquenessTheorem}.%, hence guaranteeing \eqref{Cstable} . %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
%This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\overline x}_i$ is known to have a size that grows polynomially in $m$ and $k$.
